Last glacial cycle glacier erosion potential in the Alps

### 2021-02-03 articles/erosion/alps-erosion.tex
some last changes by Ian

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large over-deepened basins. Using previous glacier
    modelling results, and empirical inferences bedrock erosion under modern
    glaciers, we compute a distribution of potential glacier erosion in the Alps
    over the last glacial cycle from 120\,000 years ago to the present.
    %
    Despite large uncertainties related to the climate history of the Alps and
    unconstrained glacier erosion processes, the resulting modelled patterns of
    glacier erosion shows persistent features. The cumulative imprint of
    the last glacial cycle shows a very strong localization of glacier erosion
    with local maxima at the mouths of major Alpine valleys and some other
    upstream sections where glaciers are modelled to have flown with the
    highest velocity. The modelled erosion rates vary significantly through the
    glacial cycle, but show paradoxically little relation to the glaciation
    extent. Phases of glacier advance and maximum extension, characterized by
    low air and glacier temperatures, see a localization of rapid erosion rates
    at low elevation, while intra-montane areas are modelled to date from phases
    of intermediate glacier cover. The modelled erosion rates peak during
    deglaciation phases, when higher air and glacier temperature enhance
    glacier dynamics.
    %
    Our results depict the Alpine glacier erosion landscape as a
    time-transgressive patchwork, with different erosional landscapes of the
    same mountain range corresponding to different glaciation stages and time
    periods, thus offering insight into its diversity.

@@

    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
      Yet, other work suggests that global weather rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
      Examination of contemporary erosion rates across different climates suggest that increased temperatures lead to higher erosion rates \citep{Koppes.etal.2015,Fernandez.etal.2016}.
      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship
      between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}

    \ian{
      Observing and quantifying the long-term erosion
      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion.
      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e. water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}

@@

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
      we examine the glaciologcial and conditions that cause erosion and discuss
      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }

@@

    \ian{Our approach does not account for the feedback of glacier erosion onto
    bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
    Neither does the approach account for downstream advection of debris \citep{Anderson.Anderson.2016}
    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
    sentence might not be that important}}

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Temporal evolution}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
  ``Annual erosion volume (m$^3$)'',
  Also, in many papers there are plots of erosion rate over time. For one plot,
  would it be good to plot erosion rate instead of volume (divide by area)?
  This way some comparison can be easier made with other literature.
  For instance, \citep{Stutenbecker.etal.2017} mentions a current erosion rate of $1.24$ mm a$^{-1}$.
  Such a scaling might be good for comparison... or not if things are way off
  and the message is much more about the relative change.
}

@@

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
      or glacier erosion  \citep[][]{Iverson.etal.2012,Ugelvig.etal.2018}}.
    The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
    sliding, not included in the model, presumably further accelerated erosion
    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
      fluvial transport of sediment, which could well increase sediment discharge
      and erosion with increased melt, so long as additional sediment remains
      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
    to accompany the  ongoing 21st-century deglaciation phase. This trend might be
    especially strong on the ice sheets of Greenland and Antarctica where
    polythermal and cold-based glaciers are warming and sliding is increasing
    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}

@@

    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre). While the validity of the model results at high elevation are
    limited by the 1-km resolution and the shallow-ice physics
    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
    of the modelled glacial cycle implies that the numerous Alpine mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
    temperate periods with limited ice-cover such as the present day \ian{
      or over interglacial periods, when glaciers are confined to their cirques
      \citep{Barr.etal.2017,Barr.etal.2019}}.

@@

    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
    erosion processes, our quantitative results are to be considered with extreme
    uncertainties. Yet to the best of our current knowledge in these respective
    areas of science, we can draw the following qualitative conclusions:
    \begin{itemize}
      \item The non-linear physics of glacier deformation and sliding, combined
        with non-linear empirical erosion laws, results in a very strong
        localization of rapid glacier erosion in regions of fast flow.
      \item Temperature-dependent creep and basal sliding provide a mechanism
        for accelerated erosion during deglaciation periods, irrespective of
        surface meltwater penetration to the glacier bed.
      \item If a non-linear glacier erosion law is used, the climate-induced
        slowdown of erosion counterbalances glacier expansion, so that
        Alpine-wide glacier erosion rates do not correlate with the ice volume.
      \item Rapid glacier erosion is restricted to low elevation during stages
        of glacier advance and maximum glaciation, but propagates up-valley to
        higher elevations during periods of glacier retreat.
      \item The diversity of the Alpine glacial erosion landscape appears to
        be the time-transgressive signature of a variety of glacial stages
        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
      \item \ian{The high spatialtemporal variability  in erosion suggests that
          the relationship between climate and glacier erosion remains greatly complex.
          As a result, a highly variable response of glacier erosion to climate forcings should be expected.}
    \end{itemize}

@@

\authorcontribution{%
    J.~Seguinot performed the computations and initially drafted the
    manuscript. I.~Delanay contributed to interpret the results in broader
    context.}

>>>>

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large over-deepened basins. Using previous glacier
    modelling results, and empirical inferences bedrock erosion under modern
    glaciers, we compute a distribution of potential glacier erosion in the Alps
    over the last glacial cycle from 120\,000 years ago to the present.
    %
    Despite large uncertainties related to the climate history of the Alps and
    unconstrained glacier erosion processes, the resulting modelled patterns of
    glacier erosion shows persistent features. The cumulative imprint of
    the last glacial cycle shows a very strong localization of glacier erosion
    with local maxima at the mouths of major Alpine valleys and some other
    upstream sections where glaciers are modelled to have flown with the
    highest velocity. The modelled erosion rates vary significantly through the
    glacial cycle, but show paradoxically little relation to the glaciation
    extent \ian{Not to fine of a point, but in the text, it is about ice volume,
      and not so much ice extent}. Phases of glacier advance and maximum extension, characterized by
    low air and glacier temperatures, see a localization of rapid erosion rates
    at low elevation, while intra-montane areas are modelled to date from phases
    of intermediate glacier cover. The modelled erosion rates peak during
    deglaciation phases, when higher air and glacier temperature enhance
    glacier dynamics. \ian{These result that the climatic conditions that
    result in glacier erosion might be quite specific. \textbf{or something like that}}
    %
    Our results depict the Alpine glacier erosion landscape as a
    time-transgressive patchwork, with different erosional landscapes of the
    same mountain range corresponding to different glaciation stages and time
    periods, thus offering insight into its diversity.

@@

    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery2009,Koppes.etal.2015,Fernandez.etal.2016}.
      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship  between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}

    \ian{
      Observing and quantifying the long-term erosion
      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion \citep{Alley.etal.2019}.
      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e. water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}

@@

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
      we examine the glaciologcial conditions that cause erosion and discuss
      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }

@@

    \ian{Our approach does not account for the feedback of glacier erosion onto
    bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
    Neither does the approach account for downstream advection of debris \citep[e.g.][]{Anderson.Anderson.2016}
    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
    sentence might not be that important}}

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Temporal evolution}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
  ``Annual erosion volume (m$^3$)''?,
  Also, in many papers there are plots of erosion rate over time. For one plot,
  would it be good to plot erosion rate instead of volume (divide by area)?
  This way some comparison can be easier made with other literature.
  For instance, \citep{Stutenbecker.etal.2017} mentions a current erosion rate of $1.24$ mm a$^{-1}$.
  Such a scaling might be good for comparison... or not if things are way off
  and the message is much more about the relative change.
}

@@

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
      or glacier erosion  \citep[][]{Iverson.etal.2012,Ugelvig.etal.2018}}.
    The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
    sliding, not included in the model, presumably further accelerated erosion
    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
      fluvial transport of sediment, which could well increase sediment discharge
      and erosion with increased melt, so long as additional sediment remains
      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
    to accompany the  ongoing 21st-century deglaciation phase. This process might be
    especially strong on the ice sheets of Greenland and Antarctica where
    polythermal and cold-based glaciers are warming and sliding is increasing
    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}

@@

    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre). While the validity of the model results at high elevation are
    limited by the 1-km resolution and the shallow-ice physics
    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
    of the modelled glacial cycle implies that the numerous Alpine mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
    temperate periods with limited ice-cover such as the present day \ian{
      or over interglacial periods, when glaciers are confined to their cirques
      \citep{Barr.etal.2017,Barr.etal.2019}}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    the mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.

@@

    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
    erosion processes, our quantitative results are to be considered with extreme
    uncertainties. Yet to the best of our current knowledge in these respective
    areas of science, we can draw the following qualitative conclusions:
    \begin{itemize}
      \item The non-linear physics of glacier deformation and sliding, combined
        with non-linear empirical erosion laws, results in a very strong
        localization of rapid glacier erosion in regions of fast flow.
      \item Temperature-dependent creep and basal sliding provide a mechanism
        for accelerated erosion during deglaciation periods, irrespective of
        surface meltwater penetration to the glacier bed.
      \item If a non-linear glacier erosion law is used, the climate-induced
        slowdown of erosion counterbalances glacier expansion, so that
        Alpine-wide glacier erosion rates do not correlate with the ice volume.
      \item Rapid glacier erosion is restricted to low elevation during stages
        of glacier advance and maximum glaciation, but propagates up-valley to
        higher elevations during periods of glacier retreat.
      \item The diversity of the Alpine glacial erosion landscape appears to
        be the time-transgressive signature of a variety of glacial stages
        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
      \item \ian{The high spatialtemporal variability  in erosion suggests that
          the relationship between climate and glacier erosion remains greatly complex.
          As a result, a highly variable response of glacier erosion to climate forcings should be expected.
        \textbf{this last point could be somewhat coupled to the first point.}}
    \end{itemize}

@@

\authorcontribution{%
    J.~Seguinot performed the computations and initially drafted the
    manuscript. I.~Delaney contributed to interpret the results in broader
    context.}
